---
title: "ggplot2 tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(leaflet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales) # install.packages("scales")
library(patchwork) # install.packages("patchwork")
```
```{r}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch( #function that tries the first argument and run the second if it gave an error
    read.csv("data/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)
```
# Data processing

```{r}
annual_esc<- esc %>%
  separate(sampleDate, c("Year","Month","Day"),sep="-",remove=F) %>%  #separate the date into its components
  mutate(Year = as.numeric(Year),
         Month= as.numeric(Month)) %>% 
  group_by(Species, SASAP.Region, Year) %>% 
  summarize(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook","Sockeye","Chum","Coho","Pink"))
  head(annual_esc)
```
# Data visualization

## First look at the data

Aes arguments are used for mapping variables to axis of the graph (incuding x,y,coor, size etc.)
The specific visual aspect is fine tuned with geom


```{r}
ggplot(annual_esc, aes(x = Species, y= escapement)) +
  geom_col(fill="blue")
```


```{r}
ggplot(annual_esc, aes(x = Species, y= escapement, fill=SASAP.Region)) +
  geom_col()
```

### Looking into the Kodiak region

```{r}
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year, y= escapement,color=Species)) +
  geom_line() + 
  geom_point() + 
  ylab("Escapement") + 
  theme_bw() +
  theme(plot.title = element_text(hjust=0.5), legend.position = "bottom") + #this needs to be after the theme_wb otherwise it gets overwritten
  ggtitle("Salmon Annual Escapement for the Kodiak Region",)

```

### other example with saving a theme as object

```{r}
my_theme<-   theme_bw() +
  theme(plot.title = element_text(hjust=0.5), 
        legend.position = "bottom"
        #,legend.title = element_blank() #to remove the legend title
  )

  
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year, y= escapement,color=Species)) +
  geom_line() + 
  geom_point() + 
  ylab("Escapement") + 
  my_theme + 
  ggtitle("Salmon Annual Escapement for the Kodiak Region",)

```

## Playing with Facet

```{r}
ggplot(annual_esc, aes(x = Year, y = escapement, color = Species)) + 
    geom_line() +
    geom_point() +
    scale_y_continuous(labels = comma) +
    facet_wrap(~SASAP.Region, scales = "free_y", ncol = 2) +
    ylab("Escapement") +
    my_theme
```

## playing with patchwork

patchwork package allows to add together multiple plots that were saved in objects

```{r}
p1<- ggplot(annual_esc, aes(x = Species, y= escapement, fill=SASAP.Region)) +
  geom_col()
p2<- annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year, y= escapement,color=Species)) +
  geom_line() + 
  geom_point() + 
  ylab("Escapement") + 
  my_theme + 
  ggtitle("Salmon Annual Escapement for the Kodiak Region",)

p1 + p2 + plot_annotation('This is a title', caption = 'made with patchwork') +
  theme(#this function uses patchwork to put the graphs together
```

