---
title: "Data cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE,message=FALSE}
library(dplyr)
library(tidyr)
```

# Read in data

We are usgin data from [Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/view/df35b.304.2)

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_original)
```



# Clean and reshape

Remove marginal sum and notes columns using `select()`

```{r}
catch_clean<-catch_original %>% 
  select(Region, Year, Chinook,Sockeye,Coho, Pink, Chum) %>% 
  mutate(Chinook=ifelse(Chinook == "I",1,Chinook)) %>% 
  mutate(Chinook=as.numeric(Chinook))
head(catch_clean)
```
## QA
Fine where as.numeric couldn't convert CHinool to a number
```{r}
i<-which(is.na(catch_clean$Chinook))
i
catch_original[i,]
```
Reshape from wide to tall format using `pivot_longer`

```{r}
catch_long <-catch_clean %>% 
pivot_longer(cols=-c(Region, Year),
              names_to= "species",
              values_to="catch") %>% 
  #the below could be more simplys done by
  #mutate(catch=catch*1000)
  rename(catch_thousands = catch) %>% 
  mutate(catch=catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)
```

Example with pivot_wider

```{r}
catch_wide<- catch_long %>% 
  pivot_wider(names_from = Year,
              values_from = catch)
head(catch_wide)
```


## Group by and summarize

```{r}
mean_region <- catch_long %>% 
  group_by(Region,species) %>% #adds the grouping information to the data
  summarise(catch_mean=mean(catch),num_obs=n())
head(mean_region)
  
```
in what year was the salmon catch the highest?
```{r}
annual_catch_sockeye <- catch_long %>% 
  filter(species=="Sockeye") %>% 
  group_by(species, Year) %>% #adds the grouping information to the data
  summarise(total_catch=sum(catch)) %>% 
  arrange(-total_catch)
annual_catch
```


# Join to regions table
```{r}
region_defs<-read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"), stringsAsFactors = F) %>% 
                         select(code, mgmtArea)
        head(reg?ion_defs)               
```
```{r}
catch_joined<-left_join(catch_long,region_defs,by=c("Region"="code"))
head(catch_joined)
```

